Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-11-07 ms/main.tex
Add yoav work in App. D

The outcomes of stochastic simulations with a structured population are shown in \autoref{fig:spatial}. The figure demonstrates that the highest cost of cooperation $c$ that permits the evolution of cooperation agrees with the conditions derived above for a model without population structure or stochasticity.
An example for stable coexistence is shown in \autoref{fig:spatial_fluctuations}.
Simulating a structured population with local selection (i.e. sites can only be settled by offspring of neighbor parents) had only a minor effect on the agreement with the derived conditions (\autoref{fig:spatial_local_selection}).

@@

We assume here that $v=1$, i.e. no oblique transmission, and therefore $\hat{p}=\tilde{p}$.
We define the frequencies of the pheno-genotypes $AM$, $BM$, $Am$, and $Bm$ as $\vec{\tilde{p}}= (\tilde{p}_1$, $\tilde{p}_2$, $\tilde{p}_3$, $\tilde{p}_4)$. 
The frequencies of the pheno-genotypes in the next generation are defined by the recursion system, 
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' = 
  & \tilde{p}_1(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_1T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_1(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_2(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)T_A \;,
\\
	% p2
  \bar{w}\tilde{p}_2' = 
  & \tilde{p}_2(\tilde{p}_2+\tilde{p}_4)(1 - (1-\alpha_1)(\tilde{p}_1 + \tilde{p}_3)T_A) + \\
  & \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_1 T_A(\tilde{p}_2 + \tilde{p}_4) - T_A(\tilde{p}_1 + \tilde{p}_3)) + \\
  & \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_1 T_B + (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_1(\tilde{p}_1 + \tilde{p}_3)(1+b-c)(1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B \;, 
\\
	% p3
  \bar{w}\tilde{p}_3' =
  & \tilde{p}_3(\tilde{p}_1+\tilde{p}_3)(1+b-c)(1 - (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_2T_B(\tilde{p}_1 + \tilde{p}_3) - T_B(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_2(\tilde{p}_2 + \tilde{p}_4)) + \\
  & \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)T_A \;, 
\\
	% p4
  \bar{w}\tilde{p}_4' =
  & \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(1 - (1-\alpha_2)(\tilde{p}_1 + \tilde{p}_3)T_A) + \\
  & \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_2T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3)) + \\
  & \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_2T_B + (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) + \\
  & \tilde{p}_3(\tilde{p}_1 + \tilde{p}_3)(1+b-c)(1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B \;,
  \end{aligned}
\end{equation}
%which we can write as 
%\begin{equation}
%\bar{w} \vec{\tilde{p}} = F(\vec{\tilde{p}}),
%\end{equation}
where $\bar{w} = 1 + (b-c)(\tilde{p}_1 + \tilde{p}_3)$ is the population mean fitness.% and $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.

The equilibrium when only allele $M$ is present is  $\vec{\tilde{p}^*} = (\tilde{p}^*, 1-\tilde{p}^*, 0, 0)$, where
\begin{equation} \label{eq:p_tilde_star_alpha_1}
\tilde{p}^*=
\frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{b(1-\alpha_1)(T_A-T_B)},
\end{equation}
which we get by setting $\alpha=\alpha_1$ and $v=1$ in \autoref{eq:general_equilibrium}.
The local stability of $\vec{\tilde{p}^*}$ to the introduction of allele $m$ is determined by the linear approximation $\vec{L}^*$ of the transformation in \autoref{eq:next_gen_p_1} near $\vec{\tilde{p}^*}$ (i.e., the Jacobian of the transformation at the equilibrium).
$\vec{L}^*$ is known to have a block structure, with the diagonal blocks occupied by the matrices $\vec{L}^*_{in}$ and $\vec{L}^*_{ex}$.
The latter is the external stability matrix: the linear approximation to the transformation near $\vec{\tilde{p}^*}$ involving only the phono-genotypes $Am$ and $Bm$.
Because we assume that $\vec{\tilde{p}^*}$ is internally stable (i.e. locally stable to small perturbations in the frequencies of $AM$ and $BM$), the stability of 
$\vec{\tilde{p}^*}$ is determined by the eigenvalues of the external stability matrix $\vec{L}^*_{ex}$, 
\begin{equation}
\mathbf{L}^*_{ex} = \frac{1}{\bar{w}^*} \begin{pmatrix}
	 X &
	 Y \\
	 Z &
	 Q 
\end{pmatrix} \;.
\end{equation}

The elements of $\mathbf{L}^*_{ex}$ are derived from \autoref{eq:next_gen_p_1},
\begin{equation} \label{eq:partial_derivatives}
  \begin{aligned}
%  X &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = b \big(\tilde{p}^*-(1 - \alpha_2) T_B (\tilde{p}^* - \tilde{p}^{*2})\big) + (1 - c) (1 + \tilde{p}^* T_B - T_B) \;,\\
  X &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = b \big(1-(1 - \alpha_2) T_B (1 - \tilde{p}^*)\big) \tilde{p}^* + (1 - c) \big(1 - T_B (1 - \tilde{p}^*)\big) \;,\\
%  Y &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = b T_A \tilde{p}^* \big(\tilde{p}^* + \alpha_2 (1 - \tilde{p}^*)\big) + \tilde{p}^* T_A \;,\\
  Y &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = b T_A \big(\tilde{p}^* + \alpha_2 (1 - \tilde{p}^*)\big) \tilde{p}^* + T_A \tilde{p}^* \;,\\
% Z &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = (1 - \tilde{p}^*) (1 - c) T_B + b \tilde{p}^* (1 - \alpha_2)(1 - \tilde{p}^*) T_B \;,\\
  Z &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = (1 - c) T_B (1 - \tilde{p}^*) + b T_B (1 - \alpha_2) \tilde{p}^*(1 - \tilde{p}^*) \;,\\
%  Q &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = 1 - \tilde{p}^*T_A+ b\tilde{p}^*\big(1 - \alpha_2 T_A (1 - \tilde{p}^*) - T_A \tilde{p}^* \big) \;.
  Q &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = 1 - T_A \tilde{p}^* + b \big(1 - T_A \tilde{p}^* - \alpha_2 T_A (1 - \tilde{p}^*) \big) \tilde{p}^* \;.
  \end{aligned}
\end{equation}

>>>>

The outcomes of stochastic simulations with a structured population are shown in \autoref{fig:spatial}. The figure demonstrates that the highest cost of cooperation $c$ that permits the evolution of cooperation agrees with the conditions derived above for a model without population structure or stochasticity.
An example for stable coexistence is shown in \autoref{fig:spatial}c.
Simulating a structured population with local selection (i.e. sites can only be settled by offspring of neighbor parents) had only a minor effect on the agreement with the derived conditions (\autoref{fig:spatial_local_selection}).

@@

We assume here that $v=1$, i.e. no oblique transmission, and therefore $\hat{p}=\tilde{p}$.
We define the frequencies of the pheno-genotypes $AM$, $BM$, $Am$, and $Bm$ as $\vec{\tilde{p}}= (\tilde{p}_1$, $\tilde{p}_2$, $\tilde{p}_3$, $\tilde{p}_4)$. 
The frequencies of the pheno-genotypes in the next generation are defined by the recursion system, where $x=\tilde{p}_1+\tilde{p}_3$ is the cooperator frequency, 
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' = 
  & \tilde{p}_1 x (1+b-c)(1 - (1-\alpha_1)(1-x)T_B) + \\
  & \tilde{p}_1(1-x)(1-c)(1-\alpha_1T_B x - T_B(1-x)) + \\
  & \tilde{p}_2 x (1+b)T_A(x + \alpha_1(1-x)) + \\
  & \tilde{p}_2(1-x)x(1-\alpha_1)T_A \;,
\\
	% p2
  \bar{w}\tilde{p}_2' = 
  & \tilde{p}_1 x (1+b-c)(1-\alpha_1)(1-x)T_B \\
  & \tilde{p}_1(1-x)(1-c)(\alpha_1 T_B + (1-\alpha_1)(1-x)T_B) + \\
  & \tilde{p}_2 x (1+b)(1-\alpha_1 T_A(1-x) - T_A x) + \\
& \tilde{p}_2(1-x)(1 - (1-\alpha_1) x T_A) \;, 
\\
	% p3
  \bar{w}\tilde{p}_3' =
  & \tilde{p}_3 x (1+b-c)(1 - (1-\alpha_2)(1-x)T_B) + \\
  & \tilde{p}_3(1-x)(1-c)(1-\alpha_2 T_B x - T_B(1-x)) + \\
  & \tilde{p}_4 x (1+b)T_A(x + \alpha_2 (1-x)) + \\
  & \tilde{p}_4(1-x) x (1-\alpha_2)T_A \;, 
\\
	% p4
  \bar{w}\tilde{p}_4' =
  & \tilde{p}_3 x (1+b-c)(1-\alpha_2)(1-x)T_B + \\
  & \tilde{p}_3(1-x)(1-c)(\alpha_2 T_B + (1-\alpha_2)(1-x)T_B) + \\
  & \tilde{p}_4 x (1+b)(1-\alpha_2T_A(1-x)-T_A x ) + \\
  & \tilde{p}_4(1-x)(1 - (1-\alpha_2) x T_A) \;,
  \end{aligned}
\end{equation}
%which we can write as 
%\begin{equation}
%\bar{w} \vec{\tilde{p}} = F(\vec{\tilde{p}}),
%\end{equation}
where $\bar{w} = 1 + (b-c) x$ is the population mean fitness.% and $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.

The equilibrium when only allele $M$ is present is  $\vec{\tilde{p}^*} = (\tilde{p}^*, 1-\tilde{p}^*, 0, 0)$, where
\begin{equation} \label{eq:p_tilde_star_alpha_1}
\tilde{p}^*=
\frac{c(1-T_B) - b \alpha_1 T_A - (T_A - T_B)}{b(1-\alpha_1)(T_A-T_B)},
\end{equation}
which we get by setting $\alpha=\alpha_1$ and $v=1$ in \autoref{eq:general_equilibrium}.
The local stability of $\vec{\tilde{p}^*}$ to the introduction of allele $m$ is determined by the linear approximation $\vec{L}^*$ of the transformation in \autoref{eq:next_gen_p_1} near $\vec{\tilde{p}^*}$ (i.e., the Jacobian of the transformation at the equilibrium).
$\vec{L}^*$ is known to have a block structure, with the diagonal blocks occupied by the matrices $\vec{L}^*_{in}$ and $\vec{L}^*_{ex}$.
The latter is the external stability matrix: the linear approximation to the transformation near $\vec{\tilde{p}^*}$ involving only the phono-genotypes $Am$ and $Bm$.
Because we assume that $\vec{\tilde{p}^*}$ is internally stable (i.e. locally stable to small perturbations in the frequencies of $AM$ and $BM$), the stability of 
$\vec{\tilde{p}^*}$ is determined by the eigenvalues of the external stability matrix $\vec{L}^*_{ex}$, 
\begin{equation}
\mathbf{L}^*_{ex} = \frac{1}{\bar{w}^*} \begin{bmatrix}
	 X &
	 Y \\
	 Z &
	 Q 
\end{bmatrix} \;.
\end{equation}

The elements of $\mathbf{L}^*_{ex}$ are derived from \autoref{eq:next_gen_p_1},
\begin{equation} \label{eq:partial_derivatives}
  \begin{aligned}
%  X &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = b \big(\tilde{p}^*-(1 - \alpha_2) T_B (\tilde{p}^* - \tilde{p}^{*2})\big) + (1 - c) (1 + \tilde{p}^* T_B - T_B) \;,\\
  X &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = 
  b \big(1-(1 - \alpha_2) T_B (1 - \tilde{p}^*)\big) \tilde{p}^* + (1 - c) \big(1 - T_B (1 - \tilde{p}^*)\big) \;,\\
%  Y &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = b T_A \tilde{p}^* \big(\tilde{p}^* + \alpha_2 (1 - \tilde{p}^*)\big) + \tilde{p}^* T_A \;,\\
  Y &= \frac{\partial\bar{w}\tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = 
  b T_A \big(\tilde{p}^* + \alpha_2 (1 - \tilde{p}^*)\big) \tilde{p}^* + T_A \tilde{p}^* \;,\\
% Z &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = (1 - \tilde{p}^*) (1 - c) T_B + b \tilde{p}^* (1 - \alpha_2)(1 - \tilde{p}^*) T_B \;,\\
  Z &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) = 
  b T_B (1 - \alpha_2) \tilde{p}^*(1 - \tilde{p}^*) + (1 - c) T_B (1 - \tilde{p}^*) \;,\\
%  Q &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = 1 - \tilde{p}^*T_A+ b\tilde{p}^*\big(1 - \alpha_2 T_A (1 - \tilde{p}^*) - T_A \tilde{p}^* \big) \;.
  Q &= \frac{\partial\bar{w}\tilde{p}_4'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) = 
  b \big(1 - T_A \tilde{p}^* - \alpha_2 T_A (1 - \tilde{p}^*) \big) \tilde{p}^* + 1 - T_A \tilde{p}^* \;.
  \end{aligned}
\end{equation}

Reorganizing these equations, we get,
\begin{equation} 
  \begin{aligned}
  	&\bar{w}^* L^*_{ex} = \\
	&\begin{bmatrix}
		(1+b \tilde{p}^* -c)(1-T_B(1-\tilde{p}^*)) + b \tilde{p}^* \alpha_2 T_B (1-\tilde{p}^*) & 
		(1+b \tilde{p}^*) T_A \tilde{p}^* + b \tilde{p}^* \alpha_2 T_A(1-\tilde{p}^*) \\
		(1+b \tilde{p}^* - c) T_B(1-\tilde{p}^*) - b \tilde{p}^* \alpha_2 T_B (1-\tilde{p}^*) &
		(1+b \tilde{p}^*) (1-T_A \tilde{p}^*) - b \tilde{p}^* \alpha_2 T_A (1-\tilde{p}^*) \\
  	\end{bmatrix} \;,
  \end{aligned}
\end{equation}
which can be rearranges as
\begin{equation} 
  \begin{aligned}
  	\bar{w}^* L^*_{ex} = 
	\Big[
	\ci 
	-\vec{T}	
	\cdot
	\vec{P}
	\Big] 
	\cdot
	\begin{bmatrix}
	1+b\tilde{p}^*-c & 0 \\ 0 & 1+b\tilde{p}^*
	\end{bmatrix}
	+ 
	\alpha_2 \cdot b \cdot \tilde{p}^* (1-\tilde{p}^*)
	\cdot \vec{T} \;,
  \end{aligned}
\end{equation}
where
\begin{equation} 
\begin{aligned}
\vec{T} &= \begin{bmatrix}
	T_B & T_A \\ -T_B & -T_A
\end{bmatrix} \;,
\vec{P} &= \begin{bmatrix}
	1-\tilde{p}^* & 0 \\ 0 & -\tilde{p}^*
\end{bmatrix} \;.
\end{aligned}
\end{equation}
